A multi-scale story of the diffusion of a new technology: the web

### 2025-01-03 paper/submission1/tranos2023.qmd
ssh problem

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  This paper maps the participation in the digital economy and its evolution in the UK over space and time. Most of the existing economic geography literature which dealt with the spatiality of the internet employed supply-side measures, such as infrastructural capacity, in order to understand the geography of the digital economy and its potential spatial economic effects. Useful as these approaches might have been, they cannot capture the micro-processes and the characteristics of the individual online behaviour. Using large volumes of archived and geolocated web content, this paper models the diffusion of web technologies over space and time in the UK. The data and geolocation strategy allow to capture these processes at a very granular spatial scale. The modelling approach, which is based on simple spatial analytical methods and on the estimation of diffusion curves at various scales, enables to depict the role of geography and other cognitive factors which drove the diffusion of web technologies. Although the focus is on a recent historical period -- 1996-2012 -- the results of the analysis depict diffusion mechanisms which can be very useful in understanding the evolutionary patterns of the adoption of other newer technologies.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

**SAY SOMETHING ABOUT DIST PAPER BY ARP, NOT SPECIFIC EXPLANATORY VARIABLES**

@@

<!-- key concepts -- copied from help.Rmd -->
Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be 
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, there is a hierarchical processes. Instead of a
horizontal distance-based diffusion mechanism, the top-down hierarchy
of urban systems shapes technological diffusion. In reality, the synthesis 
of these two processes better reflects how new technologies diffuse over 
space and time [@morrill2020spatial].

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled 'Innovation Diffusion as a Spatial Process'
[@hagerstrand1968innovation]. Hägerstrand was the first one to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified
a 'neighbourhood' mechanism of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the 'lead-lag'
spatial acceleration effect in market research [@bento2018time;
@PERES201091]. Hägerstrand is more widely known though for highlighting
the role time plays in the diffusion of innovations: an early-pioneering
period, a middle fast accelerating period and a final saturation one
[@morrill2020spatial].

@@

@grubler1990rise effectively summarises the three key spatial diffusion 
mechanisms: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy mechanisms in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and proximity. 
These are the three mechanisms that the empirical analysis in @sec-results 
investigates.

<!-- previous studies -->
The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metro areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modeled their
spatial diffusion across 402 German regions during 1970-2015. Their
key finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a
region or industry with high share of firms having a website, they are
more likely to have a website too.

At a global scale @perkins2005international explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
@perkins2011internet explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, @ding2010modeling modeled the spatial diffusion of
mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network, @lengyel2020role
analysed its adoption and the churn at a very granular spatial level.
Their results are in agreement with early theoretical and empirical
contributions reviewed here: assortativity, urban scaling and distance
are the key drivers of spatial diffusion. @bakher2013diffusion compared 
different the growth curves as a means to model diffusion of online 
shopping in Australia during the 1998-2009 period. At a global scale
@PAPAGIANNIDIS2015308 modeled the diffusion of different web
technologies and practices across countries. Interestingly, they used 
similar, but less extensive data as the ones used here. Their analysis 
illustrated how the diffusion of different web technologies and
practices follow an S-shaped pattern as well as the different diffusion
rates of the different technologies and practices.

All in all, this paper tests whether the three distinct spatial diffusion 
mechanisms shaped the diffusion of the Web in the UK at very granular
spatial scales from the Web's early stages onwards.

@@

<!-- web data -->
The counts of websites at these scales are calculated using data from the 
Internet Archive[^2] and, specifically, the JISC UK Web Domain Dataset 
[@ukwebarchive]. The Internet Archive is one of the most complete and oldest 
archive of webpages in the world operating since 1996 
[@ainsworth2011much; @holzmann2016dawn]. It is a web crawler, which discovers 
webpages by following the hyperlinks of every webpage it archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK ccTLD (.uk) from the 1996–2012 period. 
In essence, this is a long list of 2.5 billion URLs of archived webpages including 
also the archival timestamp.

@@

Instead of using the whole .uk ccTLD, this paper focuses on its commercial 
subset, the .co.uk second level domain (SLD). This choices decrease the 
heterogeneity of the web data as such commercial websites have specific 
aims: they are used to diffuse information, support online transactions 
and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as .com and these 
cases escape the data used here, such omissions should not affect the 
validity of the results given the popularity of the .uk ccTLD [@tranosuk]: 
UK consumers prefer to visit a .uk website when they are searching for 
products or services [@hope]; and canecdotal evidence indicates that during 
the first half of 2000, threec .co.uk domains were registered every minute 
[@oecd_coms]. Importantly, previous studies illustrated that .co.uk is the most 
popular UK SLD [@tranosuk].

@@

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This website 
reconstruction process lend itself to the consteruction of *website* instead 
of *webpage* density metrics. Websites tend to represent specific orgnanisations 
or entities, and, arguably, are more meaningful observational units than 
webpages, which can ignore the upstream dependency of the website they belong to. 
Based on the following example, all three webpages are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)). Only 
the overall website and not the nested webpages are considered for the counts 
as otherwise the density metrics would have been biased towards large, place 
specific, websites.

@@

\noindent What is challenging is that this aggregation approach,
which has been used elsewhere [@tranosuk; @shoreditch] may lead to websites with 
multiple postcodes. As per the above example, 
[www.website.co.uk](http://www.website.co.uk) includes two unique postcodes: 
B15 2TT and BS8 1SS. The distribution of the number of postcodes per website for 
2000 is presented in Table \ref{f2000}, which  clearly illustrates a wide range. 
At the left end of the distribution, there are websites anchored to a unique 
location (72% of all the reconstructed websites in 2000), which may represent a 
small company with a single trading location. At the right end, there are 
websites with thousands of different postcodes. Considering the time period of 
the analysis, such cases can represent directories which used to be popular 
in the pre-search engines early times of the commercial internet as well as real 
estate websites [@tranosuk].

The analysis presented here is based on two subsets of these data. Firstly, on 
websites, which only contain one unique postcode. As a robustness check, the 
analysis is replicated for an extended subset of websites, which include up to 
10 unique postcodes to capture commercial websites which point to multiple 
locations. They are geolocated by equally attaching them to these locations. 
This extended sample includes 94 percent of all the archived websites in 2000.

@@

Another data cleaning step deals with some extreme outliers. Figure 
\ref{correct} plots the website counts for the top 1000 postcodes. Some obvious 
outliers can be observed for the 2002-2006 period for a handful of postcodes, 
which can be attributed to link farms [@BL2013links]. The website counts for 
these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1 6GL, M28 2SL, DE21 7BF), which 
in 2004 or 2005 had more than 1000 websites pointing to them, were replaced 
with predicted values based on a simple panel regression model with postcode 
fixed effects and yearly dummy variables. These postcodes refer to a residential 
area with a small park in South London, residential areas outside Warwick and 
Manchester, a rural area in the south border of England and Wales and small 
business parks in Crew and Derby. Hence, it is difficult to believe that these
postcodes genuinely hosted such a large amounts of websites, which were later 
extensively declined. To put the magnitude of the data imputation into 
perspective, this process affected 6 out of the 557,808 postcodes included in 
the data. As the @sec-results illustrates, this led to a small increase of the 
model predictive capability.

@@

To create the website density metric, the yearly website counts at the LAD level 
are standardised by the number of firms in LAD to avoid biases associated with 
LAD hosting a large number of firms. Given that there is no such statistic for 
the OA, the actual OA level counts are used. However, because of the the 
consistent spatial definition of OA (they host 40-250 households),[^4] website 
counts in OA are interpreted as a density metric too.

@@

<!-- system-level analysis -->
These website density metrics are used for the three different stages of the 
analysis. Firstly, a system-level analysis explores how the diffusion of web 
technologies in the UK fits with the well-established S-curve. To do so, the 
following logistic function (Equation \ref{eq:s}) is estimated for the whole 
of the UK and for each LAD separately:

@@

<!-- ESDA -->
Next, exploratory spatial data analysis depicts whether the two main 
drivers of spatial diffusion -- namely neighbourhood and hierarchy -- underpin 
thediffusion of web technologies in the UK. To capture the former and following 
@ding2010modeling the Moran's I and the Local Index of Spatial Association 
(LISA) are estimated for the website density. To address the hierarchy mechanism, 
the Gini coefficient -- a well established metric of inequality -- is 
calculated. All the above are computed and plotted longitudinally both for the 
LAD and the OA.

<!-- RF -->
Lastly, a modelling framework is developed to test the above diffusion 
mechanisms. The overarching aim is to build a model that can test the 
relationship between website density and these mechanisms:

@@

RF have been widely employed to address regression research problems. 
@pourebrahim2019trip combined a spatial interaction modeling framework with ML 
algorithms including RF to predict commuting flows in New York City. 
@sinha2019assessing advocated for adopting spatial ensemble learning approaches, 
such as RF, to model spatial data with high autocorrelation and heterogeneity. 
@creditspatial predicted employment density in Los Angeles using spatially 
explicit RF. @guns2014recommending used RF to build a recommendation system for 
research collaborations. @ren2019predicting trained RF to predict the 
socio-economic status of cities using various online and mobility predictors. 
@tranos2023using utilised hyperlinks data and RF to make out-of-sample 
predictions of interregional trade. @zhou2023geography employed such a 
framework to assess whether key predictors of obesity differ across English 
cities.

@@

Figure \ref{s_uk} plots the cumulative adoption of the Web in total in the UK 
during the 1996-2012 period. Not surprisingly, the cumulative adoption is well 
represented by curve that resembles the figure S. The vertical 
line around year 2003 illustrates the point where the modeled cumulative 
adoption is equal to 50% of the maximum. This $t_0$ *inflection* point signals 
the maximum adoption speed and is used here to determine whether a UK LAD 
reached that inflection point earlier or later than the UK average. 
Specifically, the S curve and the inflection point are estimated for every LAD 
individually and then compared against the UK average. LAD that reached their 
inflection point earlier than the country average are labelled as *fast* and 
the rest as *slow*. Figure \ref{s_map} maps this pattern. There is a clear high 
concentration of fast adopting LAD in the South and East of England as expected. 
However, there is also such a concentration in Wales, Scotland and the North 
West of England. At a more detailed level, there are some expected 
examples of LAD with relevant industrial backgrounds delineated as fast: the 
City of London, a world-renowned cluster of finance industries [@cook2007role], 
and Reading, a town with high-tech service industries in proximity to London 
and its main airport, Heathrow [@england2005polynet]. However, there are also 
LAD which were expected to appear as fast -- e.g. Hackney in central London and 
Bristol, a well-established creative cluster [@oatley1999cultural; 
@bassett2002cultural] -- but were delineated as slow. Nevertheless, there is 
obvious spatial clustering of fast and slow LAD. Importantly, out of the ten 
fastest LAD, nine were located in the South East of England and London and one 
in Cambridge (see Table \ref{table.s.lads} in the Appendix). The above 
observations remain almost unchanged when the analysis includes websites 
with up to 10 unique postcodes -- see Figures \ref{s_uk10} and \ref{s_map10} in 
the Appendix -- advocating towards the robustness of the analysis.

@@

<!-- rankings -->
The next step is assesses the stability and volatility of the LAD in terms of 
their adoption of web technologies. As highlighted in the literature 
[@risk_perceptions], different agents have different perceptions about and 
levels of acceptance of the risks and the potential economic returns associated 
with the adoption of new technologies -- see for instance the seminal work of
@venkatesh2000theoretical. To reveal such aggregated patterns, Figure \ref{rank} 
plots the ranking of UK LAD based on website density. To decrease noise, the 
average ranking of 1996-1998 and 2010-2012 is plotted instead of the individual 
years. The colours depict the 5th and 95th percentile of the absolute 
difference of the LAD ranking between 1996-1998 and 2010-2012. While there are 
quite a few obvious cases of LAD that maintained their position between the 
beginning and the end of the study period both at the top or at the bottom of 
the hierarchy, there are also quite a few LAD that changed drastically 
their position. Some of these LAD enjoyed a process that at the first instance 
looks like *leapfrogging* since they managed to jump at the top of the hierarchy 
despite their slow start. There is extensive literature regarding the potential 
benefits of technological leapfrogging. The underpinning argument is that 
latecomers can adopt and benefit from new technologies that have been developed 
elsewhere without incurring the hefty initial R&D costs [@teece2008firm]. 
Although the leapfrogging literature does not pay much attention to cities and 
regions [@yu2018sustainability], previous research highlighted the long term 
and sustained productivity benefit of the early adopters of web technologies 
in the UK [@tranosuk]. So, although the LAD system is volatile, it is not clear 
whether the LAD with high concentration of late adopters will gain any latecomer 
benefits in a way similar to countries experiencing such technological 
leapfrogging. Still, the opportunity to catch up that @perkins2005international
illustrated at the country level, is also visible at the much more detail LAD 
scale. As previously, this is consistent with including websites with up 
to 10 postcodes (see Figure \ref{rank10}).

@@

<!-- Neighbourhood effect -->
Figures \ref{morani} and \ref{lisa} offer a first insight into whether a 
neighbourhood mechanism underpins the diffusion of the Web in the UK as they 
plot the Moran's I and the LISA maps of website density respectively for LAD 
and OA. Starting from the former, spatial autocorrelation was higher in the 
beginning of the study period and then after 2000 dropped slightly and 
stabilised around 0.2. This reflects the early concentration of high website 
density around London, which over time diffused as high-high clusters can be 
seen in other parts of the country away from London (Figure \ref{lisa}). An 
almost reverse pattern can be observed for OA. At the beginning of the study 
period Moran's I was around 0.5 and it plateaued after 2000 around 0.2. 
Because of the very small size of OA, at the early stages of the diffusion 
of web technologies their adoption was spatially scattered. This is reflected in 
the lack of any significant clusters in 1996 (Figure \ref{lisa}). Eventually, as 
the adoption rate increased, more such clusters of high website density were 
formed and this is reflected both in the Moran's I and the LISA maps in Figures 
\ref{morani} and \ref{lisa}. A similar pattern is observed for the extended 
dataset (see Figures\ref{morani10} and \ref{lisa10}). In this case, the 
magnitude of spatial autocorrelation is slightly higher illustrating a stronger 
spatial clustering mechanism. All in all, the exploratory spatial data analysis 
advocates towards an underpinning neighbourhood mechanism and the different 
scales of analysis illustrate how it evolved differently over time. 
<!-- discussion -->
Interestingly, the magnitude of Moran's I is quite similar to the ones revealed
by @ding2010modeling about mobile phone adoption in Chine albeit the much more 
detailed spatial scale of this paper.

@@

## Modelling results

>>>>

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
    Despite the broad consensus that differences in technological adoption and sophistication can lead to uneven trajectories of local development as new technologies can increase productivity and allow firms to introduce new products and services, economic geographers have shied away from exploring the diffusion of digital technologies across space and time. This paper fills this gap by exposing the roles of different spatial mechanisms in the diffusion of an intangible, digital technology at very granular geographical scales. It does so by employing a novel source of data and machine learning algorithms to model the active engagement with Web during 1996-2012.
keywords: 
  - diffusion
  - technologies
  - Web
  - digital
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

@@

<!-- key concepts -- copied from help.Rmd -->
Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be 
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovations just like pathogens spread because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, there is a hierarchical processes. According to this, instead of 
a horizontal distance-based diffusion mechanism, the top-down hierarchy
of urban systems shapes technological diffusion. In reality, the synthesis 
of these two processes better reflects how new technologies diffuse over 
space [@morrill2020spatial].

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled 'Innovation Diffusion as a Spatial Process'
[@hagerstrand1968innovation]. Hägerstrand was the first to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified
a 'neighbourhood' mechanism of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the 'lead-lag'
spatial acceleration effect in market research [@bento2018time;
@PERES201091]. Hägerstrand is more widely known though for highlighting
the role time plays in the diffusion of innovations: an early-pioneering
period, a middle fast accelerating period and a final saturation one
[@morrill2020spatial].

@@

@grubler1990rise effectively summarises the three key spatial diffusion 
mechanisms: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchical mechanism in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and proximity. 
These are the three mechanisms that the empirical analysis in @sec-results 
investigates.

<!-- previous studies -->
The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metro areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across 402 German regions during 1970-2015. Their
key finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a
region or industry with high share of firms having a website, they are
more likely to have a website too.

At a global scale @perkins2005international explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
@perkins2011internet explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the Internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, @ding2010modeling modelled the spatial diffusion of
mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network, @lengyel2020role
analysed its adoption and the churn at a very granular spatial level.
Their results were in agreement with the early theoretical and empirical
contributions reviewed here: assortativity, urban scaling and distance
are the key drivers of spatial diffusion. @bakher2013diffusion compared 
different the growth curves as a means to model diffusion of online 
shopping in Australia during the 1998-2009 period. At a global scale
@PAPAGIANNIDIS2015308 modelled the diffusion of different web
technologies and practices across countries. Using similar but less extensive 
data as the one used here, their analysis illustrated how the diffusion 
of different web technologies and practices follow an S-shaped pattern 
as well as the different diffusion rates of the different technologies 
and practices.

All in all, this paper tests whether the three distinct spatial diffusion 
mechanisms shaped the diffusion of the Web in the UK at very granular
spatial scales from the Web's early stages onwards. The next section discusses
the empirical framework.

@@

<!-- web data -->
The counts of websites at these scales are calculated using data from the 
Internet Archive[^2] and, specifically, the JISC UK Web Domain Dataset 
[@ukwebarchive]. The Internet Archive is one of the most complete and oldest 
archive of webpages in the world operating since 1996 
[@ainsworth2011much; @holzmann2016dawn]. It is a web crawler, which discovers 
webpages by following the hyperlinks of every webpage it archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK country code top level domain (ccTLD - .uk) 
from the 1996–2012 period. In essence, this is a long list of 2.5 billion URLs 
of archived webpages including also the archival timestamp.

@@

Instead of using the whole .uk ccTLD, this paper focuses on its commercial 
subset, the .co.uk second level domain (SLD). This choices decrease the 
heterogeneity of the web data as such commercial websites have specific 
aims: they are used to diffuse information, support online transactions 
and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as .com and these 
cases escape the data used here, such omissions should not affect the 
validity of the results given the popularity of the .uk ccTLD [@tranosuk]: 
UK consumers prefer to visit a .uk website when they are searching for 
products or services [@hope]; and anecdotal evidence indicates that during 
the first half of 2000, three .co.uk domains were registered every minute 
[@oecd_coms]. Importantly, previous studies illustrated that .co.uk is the 
most popular UK SLD [@tranosuk].

@@

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This website 
reconstruction process lend itself to the construction of *website* instead 
of *webpage* density metrics. Websites tend to represent specific organisations 
or entities, and, arguably, are more meaningful observational units than 
webpages, which can ignore the upstream dependency of the website they belong 
to. Based on the following example, all three webpages are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)). Only 
the overall website and not the nested webpages are considered for the counts 
as otherwise the density metrics would have been biased towards large, place 
specific, websites.

@@

\noindent What is challenging is that this aggregation approach,
which has been used elsewhere [@tranosuk; @shoreditch] may lead to websites with 
multiple postcodes. As per the above example, 
[www.website.co.uk](http://www.website.co.uk) includes two unique postcodes: 
B15 2TT and BS8 1SS. The distribution of the number of postcodes per website for 
2000 is presented in Table \ref{f2000}, which illustrates a wide range. 
At the left end of the distribution, there are websites anchored to a unique 
location (72% of all the reconstructed websites in 2000), which may represent a 
small company with a single trading location. At the right end, there are 
websites with thousands of different postcodes. Considering the time period of 
the analysis, such cases can represent directories which used to be popular 
in the pre-search engines early times of the commercial Internet as well as real 
estate websites [@tranosuk].

The analysis is firstly conducted on websites, which only contain one unique 
postcode. As a robustness check, the analysis is replicated for an extended 
subset of websites, which include up to 10 unique postcodes to capture 
commercial websites which point to multiple locations. They are geolocated 
by equally attaching them to these locations. This extended sample includes 94 
percent of all the archived websites in 2000.

@@

Another data cleaning step deals with some extreme outliers. Figure 
\ref{correct} plots the website counts for the top 1000 postcodes. Some obvious 
outliers can be observed for the 2002-2006 period for a handful of postcodes, 
which can be attributed to link farms [@BL2013links]. The website counts for 
these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1 6GL, M28 2SL, DE21 7BF), which 
in 2004 or 2005 had more than 1000 websites pointing to them, were replaced 
with predicted values based on a simple panel regression model with postcode 
fixed effects and yearly dummy variables. These postcodes refer to a residential 
area with a small park in South London, residential areas outside Warwick and 
Manchester, a rural area in the south border of England and Wales and small 
business parks in Crew and Derby. Hence, it is difficult to believe that these
postcodes genuinely hosted such a large amounts of websites, which were later 
extensively declined. To put the magnitude of the data imputation into 
perspective, this process affected 6 out of the 557,808 postcodes included in 
the data. As the @sec-results illustrates, this led to a small increase of the 
model predictive power.

@@

To create the website density metric, the yearly website counts at the LAD level 
are standardised by the number of firms in LAD to avoid biases associated with 
LAD hosting a large number of firms. Given that there is no such readily 
available statistic for the OA, the actual OA level counts are used. However, 
because of the the consistent spatial definition of OA (they host 40-250 
households),[^4] website counts in OA are interpreted as a density metric too.

@@

<!-- system-level analysis -->
These website density metrics are used for the three different stages of the 
analysis. Firstly, a system-level analysis explores how the diffusion of web 
technologies in the UK fits with the well-established S-curve of cumulative 
adoption. To do so, the following logistic function (Equation \ref{eq:s}) is 
estimated for the whole of the UK and for each LAD separately:

@@

<!-- ESDA -->
Next, exploratory spatial data analysis depicts whether the two main 
drivers of spatial diffusion -- namely neighbourhood and hierarchy -- underpin 
the diffusion of the Web in the UK. To capture the former and following 
@ding2010modeling the Moran's I and the Local Index of Spatial Association 
(LISA) are estimated for the website density. To address the hierarchy mechanism, 
the Gini coefficient -- an established metric of inequality -- is 
calculated. All the above are computed and plotted longitudinally both for the 
LAD and the OA.

<!-- RF -->
Lastly, a modelling framework is developed to test the above diffusion 
mechanisms. The overarching aim is to build a model that can test how well
these mechanisms can predict website density:

@@

RF have been widely employed to address regression research problems. 
@pourebrahim2019trip combined a spatial interaction modelling framework with ML 
algorithms including RF to predict commuting flows in New York City. 
@sinha2019assessing advocated for adopting spatial ensemble learning approaches, 
such as RF, to model spatial data with high autocorrelation and heterogeneity. 
@creditspatial predicted employment density in Los Angeles using spatially 
explicit RF. @guns2014recommending used RF to build a recommendation system for 
research collaborations. @ren2019predicting trained RF to predict the 
socio-economic status of cities using various online and mobility predictors. 
@tranos2023using utilised hyperlinks data and RF to make out-of-sample 
predictions of interregional trade. @zhou2023geography employed such a 
framework to assess whether key predictors of obesity differ across English 
cities.

@@

Figure \ref{s_uk} plots the cumulative adoption of the Web in total in the UK 
during the 1996-2012 period. Not surprisingly, the cumulative adoption is well 
represented by a curve that resembles the figure S. The vertical 
line around year 2003 illustrates the point where the modelled cumulative 
adoption is equal to 50% of the maximum. This $t_0$ *inflection* point signals 
the maximum adoption speed and is used here to determine whether a UK LAD 
reached that inflection point earlier or later than the UK average. 
Specifically, the S curve and the inflection point are estimated for every LAD 
individually and then compared against the UK average. LAD that reached their 
inflection point earlier than the country average are labelled as *fast* and 
the rest as *slow*. Figure \ref{s_map} maps this pattern. There is a clear high 
concentration of fast adopting LAD in the South and East of England as expected. 
However, there is also such a concentration in Wales, Scotland and the North 
West of England. At a more detailed level, there are some expected 
examples of LAD with relevant industrial backgrounds delineated as fast: the 
City of London, a world-renowned cluster of finance industries [@cook2007role], 
and Reading, a town with high-tech service industries in proximity to London 
and its main airport, Heathrow [@england2005polynet]. However, there are also 
LAD which were expected to appear as fast -- e.g. Hackney in central London and 
Bristol, a well-established creative cluster [@oatley1999cultural; 
@bassett2002cultural] -- but were delineated as slow. Nevertheless, there is 
obvious spatial clustering of fast and slow LAD. Importantly, out of the ten 
fastest LAD, nine were located in the South East of England and London and one 
in Cambridge (see Table \ref{table.s.lads} in the Appendix). The above 
observations remain almost unchanged when the analysis includes websites 
with up to 10 unique postcodes -- see Figures \ref{s_uk10} and \ref{s_map10} in 
the Appendix -- advocating towards the robustness of the analysis.

@@

<!-- rankings -->
The next step assesses the stability and volatility of the LAD in terms of 
their adoption of web technologies. As highlighted in the literature 
[@risk_perceptions], different agents have different perceptions about and 
levels of acceptance of the risks and the potential economic returns associated 
with the adoption of new technologies -- see for instance the seminal work of
@venkatesh2000theoretical. To reveal such aggregated patterns, Figure \ref{rank} 
plots the ranking of UK LAD based on website density. To decrease noise, the 
average ranking of 1996-1998 and 2010-2012 is plotted instead of the individual 
years. The colours depict the 5th and 95th percentile of the absolute 
difference of the LAD ranking between 1996-1998 and 2010-2012. While there are 
quite a few obvious cases of LAD that maintained their position between the 
beginning and the end of the study period both at the top or at the bottom of 
the hierarchy, there are also quite a few LAD that changed drastically 
their position. Some of these LAD enjoyed a process that at the first instance 
looks like *leapfrogging* since they managed to jump at the top of the hierarchy 
despite their slow start. There is extensive literature regarding the potential 
benefits of technological leapfrogging. The underpinning argument is that 
latecomers can adopt and benefit from new technologies that have been developed 
elsewhere without incurring the hefty initial R&D costs [@teece2008firm].
Although the leapfrogging literature does not pay much attention to cities and 
regions [@yu2018sustainability], previous research highlighted the long term 
and sustained productivity benefit of the early adopters of web technologies 
in the UK [@tranosuk]. So, although the LAD system is volatile, it is not clear 
whether the LAD with high concentration of late adopters will gain any 
latecomer-related benefits in a way similar to countries experiencing 
technological leapfrogging. Still, the opportunity to catch up that 
@perkins2005international illustrated at the country level, is also visible at 
the much more detailed LAD scale. As previously, this is consistent with 
including websites with up to 10 postcodes (see Figure \ref{rank10}).

@@

<!-- Neighbourhood effect -->
Figures \ref{morani} and \ref{lisa} offer a first insight into whether a 
neighbourhood mechanism underpins the diffusion of the Web in the UK as they 
present the Moran's I and the LISA maps of website density respectively for LAD 
and OA. Starting from the former, spatial autocorrelation was higher in the 
beginning of the study period and then after 2000 dropped slightly and 
stabilised around 0.2. This reflects the early concentration of high website 
density around London, which over time diffused as high website density clusters 
can be seen in other parts of the country away from London (Figure \ref{lisa}). 
An almost reverse pattern can be observed for OA. At the beginning of the study 
period Moran's I was around 0.5 and it plateaued after 2000 around 0.2. 
Because of the very small size of OA, at the early stages of the diffusion 
of web technologies their adoption was spatially scattered. This is reflected in 
the lack of any significant clusters in 1996 (Figure \ref{lisa}). Eventually, as 
the adoption rate increased, more such clusters of high website density were 
formed and this is reflected both in the Moran's I and the LISA maps in Figures 
\ref{morani} and \ref{lisa}. A similar pattern is observed for the extended 
dataset (see Figures\ref{morani10} and \ref{lisa10}). In this case, the 
magnitude of spatial autocorrelation is slightly higher illustrating a stronger 
spatial clustering mechanism. All in all, the exploratory spatial data analysis 
advocates towards an underpinning neighbourhood mechanism and the different 
scales of analysis illustrate how it evolved differently over time. 
<!-- discussion -->
Interestingly, the magnitude of Moran's I is similar to the ones revealed
by @ding2010modeling about mobile phone adoption in China albeit the much more 
detailed spatial scale of this paper.

@@

## Modelling outputs

The next section incorporates the above mechanisms of the spatial
diffusion of the Web into a modelling framework. The aim is to use variables 
depicting these spatial processes in order to predict the diffusion of the Web 
as a new technology in the UK over space and time and across different scales. 
Specifically, four different models are estimated. Firstly, all the data points 
for the OA and LAD are utilised in order to build two RF models and assess their 
capacity to predict the adoption of the Web at the two different scales. 
These two models will reveal the predictive capacity of the diffusion mechanisms 
and also demonstrate how the importance of such variables changes across different 
scales. The next two sets of models are again trained at these two scales: 
OA and LAD. However, instead of using all the data points, the OA and
the LAD from one of the twelve UK regions are held out for the model training.
Then, the trained model is used to predict website density for the OA or the LAD 
of the held-out region. This process takes place recursively for all twelve UK
regions. The regional differences in the predictive capacity of the different
samples will reveal how dissimilar these spatial process are across regions 
and, importantly, at different scales.

It needs to be highlighted here that the cross validation (CV) for all models
is spatially and temporally sensitive. Instead of using
10 random and space- and time-agnostic folds,the `CAST` package is employed 
which allows holding back data points from specific years and spatial units 
and use them for testing in order to estimate the model performance 
[@meyer2018improving].
